Differentiation with respect to weights for a `hyper3` object

### 2024-05-28 inst/grad_weight.Rmd
pursuant to commit 1d7c34c

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("disordR")
```

@@

$$
\frac{(wp_1)^ap_2^b}{(wp_1+p_2)^{a+b}}
$$

@@

```{r,label=definegws}
grad_weight_single <- function(H, probs, wpv, wanted){
    jj <- H[wpv]
    alpha <- drop(powers(jj))
    powers(jj) <- 1
    alpha*sum(probs[names(probs) %in% wanted])/loglik(probs,jj,log=FALSE)
}
```

@@

print(c(d1,d2))

LHS <- (d1+d2)                                          # differential from grad_weight_single()
RHS <- (loglik(probs,f(w+dw)) - loglik(probs,f(w)))/dw  # differential via numerics

print(c(lhs=LHS,rhs=RHS,diff=RHS-LHS))
```

@@

```{r slightlymorecomplicated}
H <- hyper3()
f <- function(w){
  H <- hyper3()
  H[c(p1=1 , p2=w , p3=w , p4=1)] <- 5
  H
}
w <- 1.4
dw <- 1e-6
HH <- f(w)
HH
probs <- c(p1=0.1,p2=0.2,p3=0.3,p4=0.4)
wpv <- c(p1=1 , p2=w , p3=w , p4=1)
LHS <- grad_weight_single(HH, probs, wpv, c("p2","p3"))
RHS <-  (loglik(probs,f(w=w+dw)) - loglik(probs,f(w=w)))/dw
c(RHS,LHS,LHS-RHS)
```

@@

```{r chessexample}

w <- 1.2
D <- 0.3

@@

wanted <- list("p1",NULL,NULL,"p1","p2",NULL,"p2","p3","p3")

@@

H <- f2(w=w, D=D)
H
```

@@

```{r diffwrtw}
diff <- 0
dw <- 1e-4
probs <- c(p1=0.15 , p2=0.45, p3=0.4)
B <- brackfunc(w,D)
for(i in seq_along(wanted)){
   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]])
}

@@

If we wish to differentiate with respect to `D` we need to change the `wanted` list:

```{r diffwrtD}
o <- c("p1","p2")
n <- NULL
wantedD <- list(n,o,n,o,n,n,o,n,n)

diff <- 0
dD <- 1e-5
B <- brackfunc(w,D)
for(i in seq_along(wanted)){
   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wantedD[[i]])
}

LHS <- diff  # using grad_weight_single()
RHS <- (loglik(probs,f2(w,D=D+dD)) - loglik(probs,f2(w,D)))/dD  # differential via numerics
c(LHS,RHS,LHS-RHS)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("disordR")
library("magrittr")
```

@@

$$\frac{(wp_1)^ap_2^b}{(wp_1+p_2)^{a+b}}$$

@@

```{r,label=definegws}
grad_weight_single <- function(H, probs, wpv, wanted){
  jj <- H[wpv]
  alpha <- drop(powers(jj))
  powers(jj) <- -1
  jj[setNames(1,wanted)] <- +1
  pnames(jj) <- pnames(H)
  alpha*loglik(probs,jj,log=FALSE)
  }
```

@@

LHS <- (d1+d2)                                          # differential from grad_weight_single()
RHS <- (loglik(probs,f(w+dw/2)) - loglik(probs,f(w-dw/2)))/dw  # differential via numerics
print(c(lhs=LHS,rhs=RHS,diff=RHS-LHS))
```

@@

```{r slightlymorecomplicated}
H <- hyper3()
f <- function(w){
  H <- hyper3()
  H[c(p1=1 , p2=w , p3=w )] <- -5
  H[c(p1=1)] <- 2
  H[c(p2=2)] <- 1
  H[c(p2=w,p3=w)] <- 2
  return(H)
}

w <- 1.4
dw <- 1e-6
(HH <- f(w))
probs <- c(p1=0.1,p2=0.6,p3=0.3)
d1 <- grad_weight_single(HH, probs, c(p1=1 , p2=w , p3=w), "p2")
d2 <- grad_weight_single(HH, probs, c(p1=1 , p2=w , p3=w), "p3")
d3 <- grad_weight_single(HH, probs, c(       p2=w , p3=w), "p2")
d4 <- grad_weight_single(HH, probs, c(       p2=w , p3=w), "p3")
LHS <- d1+d2+d3+d4
RHS <- (loglik(probs,f(w=w+dw/2)) - loglik(probs,f(w=w-dw/2)))/dw
  c(RHS,LHS,LHS-RHS)
```

@@

```{r chessexample}
w <- 1.2
D <- 0.3

@@

@@

  H <- f2(w=w, D=D)

```

@@

```{r diffwrtw}
probs <- c(p1=0.15 , p2=0.45, p3=0.4)
diff <- (
 grad_weight_single(H, probs, c(p1=w), "p1") +
 grad_weight_single(H, probs, c(p1=w+D,p2=1+D), "p1") +
 grad_weight_single(H, probs, c(p2=w), "p2") +
 grad_weight_single(H, probs, c(p1=1+D,p2=w+D), "p2") +
 grad_weight_single(H, probs, c(p3=w), "p3")+
 grad_weight_single(H, probs, c(p3=w,p1=1),"p3")
 )

@@

If we wish to differentiate with respect to `D` we need a different technique (TBA).
